### 2017-10-10 response/response-jde-1.tex

\item The authors interpret their results as postponement, implying they
represent a change in the timing of fertility but not its lifetime
cumulation. That is likely to be the correct interpretation, but can
they really tell? It seems to me that the current fixed effect model
does not tell us. Perhaps they could include 1-2 lags in the crop shock
variable to shed some light on it.

[RESPONSE:]

It is correct that the current fixed effects model cannot tell
us whether we observe postponement of fertility or a permanent reduction.
We focus on postponement for two reasons.

First, we consider the ability to actively postpone fertility as a 
necessary---but not sufficient---condition for permanent fertility reduction.
In other words, we see postponement as a weaker claim than permanent 
fertility reduction.
If we cannot show a conscious decision to delay fertility under the
right circumstances, it is hard to argue that we can convincingly
show find evidence of conscious, permanent reduction in fertility.
Note that this does not follow the original demography
literature, but is in line with the more recent research
\citep[See discussion in][]{Cinnirella2017}.

Second, from a theoretical standpoint it is not clear that 
shocks---in and off themselves---should substantially affect completed 
fertility. 
We now discuss evidence from Guatemala that the overall \emph{risk} of 
experiencing a shock is likely to affect fertility, but that only
shocks that occur late in a woman's reproductive life cycle is
likely to affect completed fertility (because the opportunities to
make up for the ``missed'' birth are fewer and because the marginal
benefit of a child is smaller the more children you already have and
the older you are).
Hence, an important distinction is that long term fertility is likely
affected by the overall \emph{risk}, whereas the exact timing of
fertility is driven by a combination of risk and shocks, but probably
mostly shocks.

% The longitudinal nature of the data is important for our ability
% to control for unobservable village, household, and individual 
% characteristics, but the trade-off is that the data only cover 
% 32 months and that is too short to analyze long-term fertility
% outcomes.
That said, we did follow your suggestion of adding an additional 
lag to the model and discuss the results in a new Appendix section.
In principle, if shocks lead to a postponement of fertility,
crop losses that occurred further in the past should be
associated with an \emph{increased} likelihood of current pregnancy 
or birth (an initial reduction in fertility should be followed
by a compensatory increase later on), 
whereas if shocks lead to a permanent reduction in fertility we
should not see any compensatory effect.

Although the results are potentially interesting in themselves,
because of data limitation the approach unfortunately does not  
provide a clean test for whether households postpone or reduce their 
fertility in response to shocks.
First, given the short duration between surveys and the lack of 
information about when the shocks occurred, it is possible and
even likely that the effects of a crop loss last for more than one period.
Hence, if we find a negative effect of more distant crop losses
this may be the result of the negative effects of the crop losses
still affecting the household.
Second, we lose 1/4 of the original number of observations each 
time we include an additional lag, which leaves us with less variation 
in crop loss because the first round had a higher proportion of crop 
losses than the subsequent rounds.

Even if crop losses' impact are limited to only one period, there 
are two additional concerns.
First, the compensatory effect may be spread over multiple 
periods because of the uncertainty associated with waiting time
to conception, i.e. a couple may be trying to conceive in the
period after the crop loss but fail.
This makes a positive compensatory effect harder to 
identify.
Second, it is possible that households may respond differently to a 
shock if they are hit repeatedly.
There are two competing factors at work.
On one hand, repeated shocks may exacerbate the negative impact on the
household, making it \emph{more} likely that fertility is postponed.
On the other hand, even longer spacing between births can be costly 
because economies of scale are harder to realize and because the 
mother may have to  be out of the labor force for longer if she 
already have young children to care for, which will make it
\emph{less} likely that fertility is postponed.
To capture the idea that repeated shocks have an additional
impact we therefore also include the interaction between 
crop losses.

With the data limitations and additional concerns,
it is difficult to draw strong predictions on what the coefficients
should be (except that the initial impact should lead to postponement 
of fertility).
Including only the additional lag of crop loss it appears that 
there is a weak positive association between crop loss in the prior
period and the likelihood of pregnancy and birth, although the 
coefficients are small and statistically insignificant.
There is still a negative effect of the immediate crop loss
variable, although these are also not statistically significant.
These results are broadly consistent with a compensatory response
where crop loss longer occur leads to higher fertility now.
However, if we include the interaction between crop losses we 
instead find negative association for both immediate and lagged 
crop loss with a positive interaction coefficient.
Even with the positive interaction coefficient the estimated 
impact of crop losses in the two prior periods is negative.
For contraceptive use we see similar patterns (with the signs 
reversed).

Although these results should be interpreted with caution, they
are not inconsistent with a compensatory/postponement effect, 
as long as some crop losses' impact last for more than one
period after its occurrence and/or the compensatory effect is
spread over more than one period.
In either specification the coefficients on lagged crop loss
are small relative to the immediate impact.
In sum, we cannot establish with any degree of confidence from
these results whether the response to shocks lead to only postponement 
of shocks or a permanent reduction.
Given the findings from Guatemala that shocks only 
have a negative impact on completed fertility if they occur
late in a woman's reproductive life and that completed 
fertility is mainly driven by the \emph{risk} of shocks,
it is most likely that what we observe are, indeed, postponement
rather than a reduction in completed fertility \citep{Portner2014}.

@@

Although we cannot measure directly measure stress, we do have information 
on illnesses and their durations.
If stress is an important factor in reporting abstinence, we should
expect the same to be the case with illness, especially those that 
prevent a woman from working.
Hence, although we did not explicitly cover this possibility in the original 
version, coital frequency may also decrease because of illness.
We have added that possibility to the text.

% We would expect illness to go up both because of the direct effect
% on the household and from increased stress.

We show the effect of crop loss on illness in Table 5.
Crop loss is  not associated with substantial enough changes in BMI or 
self-reported illness for the women in the sample to explain our fertility results.
Furthermore, we ran regressions (not included in the paper) that show that there 
is only a very weak positive association between having been ill for two or more 
days in the last 6 months and the likelihood of reporting using abstinence 
as a contraceptive strategy.
This relationship becomes only marginally stronger, and never anywhere near
statistically significant, if defining illness as not being able to
perform usual activities for 7 days or more.
Going above 7 days makes the relationship negative, most likely because
very few women report being unable to perform regular activities for that
long.

To further investigate the role of abstinence we added a new
table to the Appendix, which further subdivides the results in Table 6.
Appendix Table A-8 shows, not surprisingly, that abstinence is the most 
effective traditional contraception, although only in combination 
with a crop loss as in the table in the main text.
A major caveat is that only 19 women report using abstinence only
across the three waves used for these regressions.
Modern contraceptive also are mainly effective in combination with
a crop loss, and the direct effect of using modern contraceptives
is not statistically significant.
We discuss these results toward the end of the section on 
``Conscious Decision or Unintended Consequences?''.

### 2017-10-10 paper/incomeShocks-jde-r1.tex
Respond to R1 on traditional contraceptives

[TK This last part is crap!]

Finally, the Appendix discuss in detail whether the results 
vary by age group, wealth, and education levels.
None of these lead to strong conclusions on how crop loss 
affect fertility decision by household and individual
characteristics.
The Appendix also discuss how including additional lags
of crop loss affects the results.
In principle, if shocks lead to a postponement of fertility we 
should see longer lags be associated with higher fertility, 
whereas if shocks lead to a permanent reduction in fertility we
should not see any compensatory effect.
There are, however, a number of issues with this approach.
First, it is possible that the effect of a crop loss is felt
in subsequent period and not only in the period in which it
was observed.
Second, longer spacing between births can be costly because economies
of scale are harder to realize and because the mother may have to 
be out of the labor force for longer if she already have young
children to care for.
This implies that a household that are hit twice in a row by
a crop loss is likely to behave differently from a household that
is hit once.
Both of these factors make it difficult to draw strong 
predictions on what the effects should be (except that the
initial impact should lead to postponement of fertility).

@@

% [Lead with this or the next sentence?]
% Our result that people are able to postpone fertility using mainly 
% traditional contraceptive do, however, have implications for our 
% understanding of long-term fertility.
% [Alternative lead]
What implications do our results have for our understanding of 
long-term fertility?
% What we do not know is whether the psychic cost of using traditional 
% contraceptives is so high that although households are willing to use 
% them to deal with short-term shocks, they become difficult to use for 
% long-run moderation of fertility.
The economic literature on fertility transition in Europe and the US 
show a substantial decline in fertility and control of the timing of 
fertility before the introduction of modern contraceptives, although 
the methods used are generally not well documented.%
\footnote{
See, for example, \citet{Guinnane2011} on fertility reduction and 
\citet{Cinnirella2017} on spacing of birth.
On contraceptive methods used see, for example, \citet{Michael1976},
\citet{David1986}, and \citet{Santow1995}.
}
Furthermore, studies of the introduction of modern family planning, 
using individual level data, rarely find reductions in fertility larger 
than one child.%
\footnote{
See, for example, \citet{Portner2014a} 
and \citet{Portner2018} and the literature surveyed therein.
}
Both set of results indicate that households are able to 
limit long-term fertility using traditional contraceptives.
Our individual level data allow us to directly link the use of
traditional contraceptives and fertility outcomes.
We argue that, contrary to common perception, traditional
contraceptive methods have a high degree of efficacy when the
incentives to prevent pregnancy or birth are sufficiently high.
Because of the small data set we cannot, however, differentiate between
the different traditional methods, although---contrary to Europe 
and the US---withdrawal plays little or no role in this area.

@@

In this section we discuss how including additional lags
of crop loss affects the results.
In principle, if shocks lead to a postponement of fertility we 
should see longer lags be associated with higher fertility, 
whereas if shocks lead to a permanent reduction in fertility we
should not see any compensatory effect.
There are, however, a number of issues with this approach.
First, it is possible that the effect of a crop loss is felt
in subsequent periods and not only in the period in which it
was observed.
Second, we lose 1/4 of observations each time we include an 
additional lag, which leaves us with less variation because
the first round had a higher proportion of crop losses than
the subsequent rounds.
This is especially a concern for contraception use because
of the higher likelihood of measurement error discussed in the
paper.

Furthermore, even if shocks only had an impact in one period, 
there are two competing factors at work here.
First, a negative shocks will make people more likely to postpone 
their pregnancy/birth because of the factors discussed in the
paper.
Second, longer spacing between births can be costly because economies
of scale are harder to realize and because the mother may have to 
be out of the labor force for longer if she already have young
children to care for.
This implies that a household that are hit twice in a row by
a crop loss is likely to behave differently from a household that
is hit once.
Both of these factors make it difficult to draw strong 
predictions on what the effects should be (except that the
initial impact should lead to postponement of fertility).
We have therefore also added an interaction between crop loss for 
different periods with the idea that there is a cost to continually 
postponing births.

@@

Tables \ref{tab:postpone_birth} and \ref{tab:postpone_contraceptive}
show three models, the main results presented in the paper, one
where an additional lag is included, and one where the interaction
between crop loss and the additional lag in crop loss are included.
Including only the additional lag of crop loss it appears that 
there is a weak positive association between crop loss in the prior
period and the likelihood of pregnancy and birth, although the 
coefficients are small and statistically insignificant.
There is still a negative effect of the immediate crop loss
variable, although these are also not statistically significant.
If we include the interaction between crop losses we instead
find negative association for both immediate and lagged crop
loss with a positive interaction coefficient.
For contraceptive use we see a similar pattern (of course,
with the signs reversed).

Although these results should be interpreted with caution the
results appear most in line with the following conclusions.
First, crop losses tend to affect not only this period but also 
(weakly) the period after as indicated by the negative coefficient 
on the lagged crop loss.
Second, experiencing two crop losses in a row makes it \emph{less}
likely for people to postpone fertility (although the effect 
of the crop losses is still negative), most likely because of
the cost associated with longer spacing.

>>>>

Finally, in the Appendix we discuss in detail whether the results 
vary by age group, wealth, and education levels.
None of these lead to strong conclusions on how crop loss 
affect fertility decision by household and individual
characteristics.
Similarly, we discuss in the Appendix whether we can 
distinguish between a postponement of fertility and a
permanent reduction as a result of crop losses.
Unfortunately, because of data limitations this is difficult
to show with any degree of confidence, but we believe that 
postponement of fertility is more probable than a permanent 
reduction, which is more likely to be affected by the risk 
of a shock \citep{Portner2014}.

% The Appendix also discuss how including additional lags
% of crop loss affects the results.
% In principle, if shocks lead to a postponement of fertility we 
% should see longer lags be associated with higher fertility, 
% whereas if shocks lead to a permanent reduction in fertility we
% should not see any compensatory effect.
% There are, however, a number of issues with this approach.
% First, it is possible that the effect of a crop loss is felt
% in subsequent period and not only in the period in which it
% was observed.
% Second, longer spacing between births can be costly because economies
% of scale are harder to realize and because the mother may have to 
% be out of the labor force for longer if she already have young
% children to care for.
% This implies that a household that are hit twice in a row by
% a crop loss is likely to behave differently from a household that
% is hit once.
% Both of these factors make it difficult to draw strong 
% predictions on what the effects should be (except that the
% initial impact should lead to postponement of fertility).

@@

% [Lead with this or the next sentence?]
% Our result that people are able to postpone fertility using mainly 
% traditional contraceptive do, however, have implications for our 
% understanding of long-term fertility.
% [Alternative lead]
What implications, if any, do our results then have for our understanding 
of long-term fertility?
% What we do not know is whether the psychic cost of using traditional 
% contraceptives is so high that although households are willing to use 
% them to deal with short-term shocks, they become difficult to use for 
% long-run moderation of fertility.
The economic literature on fertility transition in Europe and the US 
show a substantial decline in fertility and control of the timing of 
fertility before the introduction of modern contraceptives, although 
the methods used are generally not well documented.%
\footnote{
See, for example, \citet{Guinnane2011} on fertility reduction and 
\citet{Cinnirella2017} on spacing of birth.
On contraceptive methods used see, for example, \citet{Michael1976},
\citet{David1986}, and \citet{Santow1995}.
}
Furthermore, studies of the introduction of modern family planning, 
using individual level data, rarely find reductions in fertility larger 
than one child.%
\footnote{
See, for example, \citet{Portner2014a} 
and \citet{Portner2018} and the literature surveyed therein.
}
Both set of results indicate that households are able to 
limit long-term fertility using traditional contraceptives.
Our individual level data allow us to directly link the use of
traditional contraceptives and fertility outcomes.
We argue that, contrary to common perception, traditional
contraceptive methods have a high degree of efficacy when the
incentives to prevent pregnancy or birth are sufficiently high.
Because of the small data set we cannot, however, differentiate between
the different traditional methods, although---contrary to Europe 
and the US---withdrawal plays little or no role in this area.

@@

When discussing our results in the paper, we treat the 
lower likelihoods of pregnancy and birth as a postponement 
of fertility.
In this section, to try to disentangle whether the lower 
likelihoods reflect a postponement or a more permanent reduction 
in fertility, we discuss how including additional lags
of crop loss affects the results.%
\footnote{
We are grateful to an anonymous reviewer for suggesting 
this approach.
}

In principle, if shocks lead to a postponement of fertility,
crop losses that occurred further in the past should be
associated with an \emph{increased} likelihood of current pregnancy 
or birth (an initial reduction in fertility should be followed
by a compensatory increase later on), 
whereas if shocks lead to a permanent reduction in fertility we
should not see any compensatory effect.

Unfortunately, because of data limitation, this approach does not 
provide a clean test for whether households 
postpone or reduce their fertility in response to shocks.
First, given the short duration between surveys and the lack of 
information about when the shocks occurred, it is possible and
even likely that the effects of a crop loss last for more than one period.
Hence, if we find a negative effect of more distant crop losses
this may be the result of the negative effects of the crop losses
still affecting the household.
Second, we lose 1/4 of the original number of observations each 
time we include an additional lag, which leaves us with less variation 
in crop loss because the first round had a higher proportion of crop 
losses than the subsequent rounds.

Even if crop losses' impact are limited to only one period, there 
are two additional concerns.
First, the compensatory effect may be spread over multiple 
periods because of the uncertainty associated with waiting time
to conception, i.e. a couple may be trying to conceive in the
period after the crop loss but fail.
This makes a positive compensatory effect harder to 
identify.
Second, it is possible that households may respond differently to a 
shock if they are hit repeatedly.
There are two competing factors at work.
On one hand, repeated shocks may exacerbate the negative impact on the
household, making it \emph{more} likely that fertility is postponed.
On the other hand, even longer spacing between births can be costly 
because economies of scale are harder to realize and because the 
mother may have to  be out of the labor force for longer if she 
already have young children to care for, which will make it
\emph{less} likely that fertility is postponed.
To capture the impact of repeated shocks we therefore also 
include the interaction between crop losses.

@@

With the data limitations and additional concerns,
it is difficult to draw strong predictions on what the coefficients
should be (except that the initial impact should lead to postponement 
of fertility).
Tables \ref{tab:postpone_birth} and \ref{tab:postpone_contraceptive}
show three models, the main results presented in the paper, one
where an additional lag is included, and one where the interaction
between crop loss and the additional lag in crop loss are included.

Including only the additional lag of crop loss it appears that 
there is a weak positive association between crop loss in the prior
period and the likelihood of pregnancy and birth, although the 
coefficients are small and statistically insignificant.
There is still a negative effect of the immediate crop loss
variable, although these are also not statistically significant.
These results are broadly consistent with a compensatory response
where crop loss longer occur leads to higher fertility now.
However, if we include the interaction between crop losses we 
instead find negative association for both immediate and lagged 
crop loss with a positive interaction coefficient.
Even with the positive interaction coefficient the estimated 
impact of crop losses in the two prior periods is negative.
For contraceptive use we see similar patterns (with the signs 
reversed).

Although these results should be interpreted with caution, they
are not inconsistent with a compensatory/postponement effect, 
as long as some crop losses' impact last for more than one
period after its occurrence and/or the compensatory effect is
spread over more than one period.
In either specification the coefficients on lagged crop loss
are small relative to the immediate impact.
In sum, we cannot establish with any degree of confidence from
these results whether the response to shocks lead to only postponement 
of shocks or a permanent reduction.
Given the findings from Guatemala that shocks only 
have a negative impact on completed fertility if they occur
late in a woman's reproductive life and that completed 
fertility is mainly driven by the \emph{risk} of shocks,
it is most likely that what we observe are, indeed, postponement
rather than a reduction in completed fertility \citep{Portner2014}.

% First, crop losses tend to affect not only this period but also 
% (weakly) the period after as indicated by the negative coefficient 
% on the lagged crop loss.
% Second, experiencing two crop losses in a row makes it \emph{less}
% likely for people to postpone fertility (although the effect 
% of the crop losses is still negative), possibly because of
% the cost associated with longer spacing.
